---
title: "DV_final_Project"
author: "Richard Li, Xiaokuan Ye, and James Mao"
date: "2024-04-23"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r , echo=TRUE}
transportation_data <- read.csv("~/Desktop/Data Visualization/Final Project/THT_Data_5082.csv", header=FALSE)
#transportation_data <- read.csv("~/Downloads/THT_Data_5082.csv", header=FALSE)
```

```{r , echo=TRUE}
#install.packages("leaflet")
#install.packages("igraph")
#install.packages("ggraph")
#install.packages("tidyr")
library(leaflet)
library(dplyr)
library(ggplot2)
library(maps)
library(wordcloud)
library(tm)
library(igraph)
library(ggraph)
library(shiny)
library(tidyr)
```

##1.static images based on ggplot2

Average Traffic Fatalities by Commute Mode:we are comparing average road traffic fatalities for auto, bicycle, and pedestria

```{r, echo=TRUE}
# Means of fatalities
fatalities_data <- transportation_data %>%
  summarise(
    Auto_Fatalities = mean(as.numeric(V24), na.rm = TRUE),
    Bicycle_Fatalities = mean(as.numeric(V26), na.rm = TRUE),
    Pedestrian_Fatalities = mean(as.numeric(V28), na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "Mode", values_to = "Fatalities")

# The Plot
ggplot(fatalities_data, aes(x = Mode, y = Fatalities, fill = Mode)) +
  geom_col() +
  labs(x = "Mode of Transport", y = "Average Fatalities per 100,000 Residents") +
  ggtitle("Comparison of Average Traffic Fatalities by Commute Mode") +
  theme_minimal()
# # Means of fatalities
# fatalities_data <- transportation_data %>%
#   summarise(
#     Auto_Fatalities = mean(as.numeric(V24), na.rm = TRUE),
#     Bicycle_Fatalities = mean(as.numeric(V26), na.rm = TRUE),
#     Pedestrian_Fatalities = mean(as.numeric(V28), na.rm = TRUE)
#   ) %>%
#   pivot_longer(cols = everything(), names_to = "Mode", values_to = "Fatalities")
# 
# # The Plot
# ggplot(fatalities_data, aes(x = Mode, y = Fatalities, fill = Mode)) +
#   geom_col() +
#   labs(x = "Mode of Transport", y = "Average Fatalities per 100,000 Residents") +
#   ggtitle("Comparison of Average Traffic Fatalities by Commute Mode") +
#   theme_minimal()

```

Seat Belt Use Score Across Different States: we are visualizing the distribution of seat belt use scores across states. - Apparently Washington, Oregon, California, Texas and Minnesota are the top 5 states that uses seat belt, which could be very possibly associated with lower fatality rates. On the other hand, South Dakota, New Hampshire, Arkansas, Massachusetts, and Montana are the bottom 5 states that uses seat belt while driving.

States like Washington, Oregon, and California, known for their progressive attitudes towards safety and stronger enforcement of traffic laws, likely exhibit higher seat belt usage. These states often have dense urban populations, well-maintained road networks, and significant public awareness campaigns promoting safety measures. In contrast, states with lower seat belt usage such as South Dakota and Montana may have more rural populations, longer distances between destinations, and a culture that places less emphasis on strict adherence to safety regulations.

```{r, echo=TRUE}

transportation_data <- transportation_data %>%
  mutate(V37 = as.numeric(as.character(V37))) # Convert from factor to numeric if necessary

sum(is.na(transportation_data$V37))

# Creating a boxplot
ggplot(transportation_data, aes(x = reorder(V2, V37, FUN = median), y = V37, fill = V2)) + 
  geom_boxplot() +
  scale_fill_viridis_d() +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "none") +  
  labs(x = "State", y = "Seat Belt Use: Score", title = "Distribution of Seat Belt Use Scores Across States") +
  coord_flip()  

```

As expected, the seat belt use and fatality rate is negatively correlated: the trendline fit with the pattern of the scatters, higher use of seat belt is associated in lower expected fatality rates.

```{r, echo=TRUE}
# We are checking and ensuring both columns are numeric
transportation_data$V37 <- as.numeric(as.character(transportation_data$V37))
transportation_data$V24 <- as.numeric(as.character(transportation_data$V24))

# Plotting a scatter plot, also with a fitted line
ggplot(transportation_data, aes(x = V37, y = V24)) +
  geom_point(aes(color = V2), size = 3) +  
  geom_smooth(method = "lm", se = FALSE, color = "black") +  
  scale_color_viridis_d() +  
  theme_minimal() +
  labs(x = "Seat Belt Use: Score", 
       y = "Road Traffic Fatalities per 100,000 Residents - Auto", 
       title = "Relationship Between Seat Belt Use Scores and Auto Fatalities") +
  theme(legend.position = "none")  
```

##2. maps using geospatial data

Physical activity score: measure of physical activity associated with transportation in each state. 'Commute Mode Share': the percentage of people using active modes of transportation such as walking, biking, or public transit for their daily commute. States with higher percentages of people using these modes might have higher physical activity scores, which leads to a potentially better health outcomes.

```{r, echo=TRUE}
# transportation_data$V20 <- as.numeric(as.character(transportation_data$V20))
# 
# sum(is.na(transportation_data$V20))
# 
# transportation_data$state <- tolower(as.character(transportation_data$V2))
# us_map <- map_data("state")
# map_data <- merge(us_map, transportation_data, by.x = "region", by.y = "state")
# 
# # Plotting the map
# ggplot(data = map_data, aes(x = long, y = lat, group = group, fill = V20)) + 
#   geom_polygon(color = "white") + 
#   geom_path() + 
#   coord_fixed(1.3) + 
#   scale_fill_viridis_c(name = "Physical Activity Score") +  
#   theme_void() +
#   labs(title = "Physical Activity by State", subtitle = "Based on Commute Mode Share")

#------------------------------------------------------------------------------
transportation_data$V20 <- as.numeric(as.character(transportation_data$V20))

transportation_data$state_code <- state.abb[match(transportation_data$V2, state.name)]

us_map <- map_data("state")
map_data <- merge(us_map, transportation_data, by.x = "region", by.y = "V3")

state_centers <- map_data %>%
  group_by(region) %>%
  summarise(center_long = mean(long, na.rm = TRUE),
            center_lat = mean(lat, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(transportation_data, by = c("region" = "V3"))

# map with state codes
gg <- ggplot(data = map_data, aes(x = long, y = lat, group = group, fill = V20)) +
  geom_polygon(color = "white") +
  geom_path() +
  scale_fill_viridis_c(name = "Physical Activity Score") +
  theme_void() +
  labs(title = "Physical Activity by State", subtitle = "Based on Commute Mode Share")

gg <- gg + geom_text(data = state_centers, aes(x = center_long, y = center_lat, label = state_code, group = NULL), size = 3, hjust = 0.5, vjust = 0.5)

gg + coord_fixed(1.3)

#A leaflet interactive map with activity scores
transportation_data$V20 <- as.numeric(as.character(transportation_data$V20))
transportation_data$state_code <- state.abb[match(transportation_data$V2, state.name)]
us_map <- map_data("state")
us_map$region <- tolower(us_map$region)

map_data <- merge(us_map, transportation_data, by.x = "region", by.y = "V3", all.x = TRUE)

state_centers <- map_data %>%
  group_by(region) %>%
  summarise(center_long = mean(long, na.rm = TRUE),
            center_lat = mean(lat, na.rm = TRUE),
            activity_score = mean(V20, na.rm = TRUE),
            state_name = first(V2))  

pal <- colorNumeric(palette = "viridis", domain = state_centers$activity_score)

# Creating the interactive map
leaflet(state_centers) %>%
  addTiles() %>%
  addCircles(lng = ~center_long, lat = ~center_lat, weight = 1,
             radius = 50000, color = ~pal(activity_score),
             fillOpacity = 0.8, popup = ~paste(state_name, "<br>Activity Score: ", activity_score)) %>%
  addLegend("bottomright", pal = pal, values = ~activity_score,
            title = "Activity Score", opacity = 0.7)
```

##4 network analysis network 
Represents states with similar commuting patterns. - The dense cluster of nodes in the center with many overlapping edges suggests that many states have similar commuting patterns according to your similarity criteria. This could mean that most states share common percentages of commuters using auto or transit modes - Outliers: States like New York seem to be outliers with very few connections or possibly none at all. This implies that New York's commuting patterns are significantly different from most other states. Given New York's extensive public transportation system, this might reflect a higher use of transit compared to other states.

```{r, echo=TRUE}

transportation_data$V4 <- as.numeric(as.character(transportation_data$V4))
transportation_data$V6 <- as.numeric(as.character(transportation_data$V6))

# Here calculates the similarity matrices for auto and transit mode shares
auto_similarity <- outer(transportation_data$V4, transportation_data$V4, FUN = function(x, y) abs(x - y) < 0.1 * x)
transit_similarity <- outer(transportation_data$V6, transportation_data$V6, FUN = function(x, y) abs(x - y) < 0.1 * x)

combined_similarity <- auto_similarity | transit_similarity

graph <- graph_from_adjacency_matrix(combined_similarity, mode = "undirected", diag = FALSE)
V(graph)$name <- transportation_data$V2

# plotting the network graph

ggraph(graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = 0.5), show.legend = FALSE) +
  geom_node_point(color = "darkblue", size = 3) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_graph() +
  labs(title = "Network of State Similarities in Commuting Patterns")
```

```{r, echo=TRUE}
# Setting the first row as column names and removing the first row from the dataframe
transportation_data <- read.csv("~/Desktop/Data Visualization/Final Project/THT_Data_5082.csv", header=TRUE)
# colnames(transportation_data) <- as.character(unlist(transportation_data[1, ]))
# transportation_data <- transportation_data[-1, ]

```

##Shiny Search Bars 
What we want to achieve here are 2 search boxes with these features and functionalities: - state -type of transportation -\> fatality score, state map

```{r, echo=TRUE}

# The Shiny UI
ui <- fluidPage(
    titlePanel("Transportation Fatality Search"),
    sidebarLayout(
        sidebarPanel(
            selectInput("state", "Select a State:", choices = c(
                "Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado",
                "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho",
                "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana",
                "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota",
                "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada",
                "New Hampshire", "New Jersey", "New Mexico", "New York",
                "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon",
                "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota",
                "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington",
                "West Virginia", "Wisconsin", "Wyoming")),
            selectInput("transportType", "Select Type of Transportation:", 
                        choices = c("Auto" = "Road.Traffic.Fatalities.per.100.000.Residents...Auto..Score", 
                                    "Bicycle" = "Road.Traffic.Fatalities.per.100.000.Residents...Bicycle..Score", 
                                    "Pedestrian" = "Road.Traffic.Fatalities.per.100.000.Residents...Pedestrian..Score")),
            actionButton("search", "Search")
        ),
        mainPanel(
            textOutput("result"),
            leafletOutput("stateMap", height = 250)  # Placeholder for map output
        )
    )
)

# The Shiny server
server <- function(input, output) {
    observeEvent(input$search, {
        
        fatality_score <- transportation_data[transportation_data$state == input$state, input$transportType]
        
         # Output the result
        output$result <- renderText({
            if (!is.na(fatality_score) && length(fatality_score) > 0) {
                paste("The fatality score for", input$transportType, "transportation in", input$state, "is", fatality_score,
                      "\nState: ", input$state,
                      "\nTransportation Type: ", input$transportType,
                      "\nFatality Score: ", fatality_score)
            } else {
                paste("Data not available for the selected state and transportation type.",
                      "\nState: ", input$state,
                      "\nTransportation Type: ", input$transportType)
            }
        })
        
transportation_data[['Road.Traffic.Fatalities.per.100.000.Residents...Auto..Score']] <- as.numeric(as.character(transportation_data[['Road.Traffic.Fatalities.per.100.000.Residents...Auto..Score']]))
transportation_data[['Road.Traffic.Fatalities.per.100.000.Residents...Bicycle..Score']] <- as.numeric(as.character(transportation_data[['Road.Traffic.Fatalities.per.100.000.Residents...Bicycle..Score']]))
transportation_data[['Road.Traffic.Fatalities.per.100.000.Residents...Pedestrian..Score']] <- as.numeric(as.character(transportation_data[['Road.Traffic.Fatalities.per.100.000.Residents...Pedestrian..Score']]))
#print('here')
transportation_data$state_code <- state.abb[match(transportation_data$state, state.name)]

us_map <- map_data("state")

us_map$region <- tolower(us_map$region)

map_data <- merge(us_map, transportation_data, by.x = "region", by.y = "region", all.x = TRUE)

#state_centers
# here we calculated the center longitude and latitude for each state
map_data$`Road.Traffic.Fatalities.per.100.000.Residents...Auto..Score`= as.numeric(map_data$`Road.Traffic.Fatalities.per.100.000.Residents...Auto..Score`)
map_data$`Road.Traffic.Fatalities.per.100.000.Residents...Bicycle..Score`= as.numeric(map_data$`Road.Traffic.Fatalities.per.100.000.Residents...Bicycle..Score`)
map_data$`Road.Traffic.Fatalities.per.100.000.Residents...Pedestrian..Score`= as.numeric(map_data$`Road.Traffic.Fatalities.per.100.000.Residents...Pedestrian..Score`)
state_centers <- map_data %>%
  # mutate(
  #   `Road Traffic Fatalities per 100,000 Residents - Auto: Score` = as.numeric(`Road Traffic Fatalities per 100,000 Residents - Auto: Score`),
  #   `Road Traffic Fatalities per 100,000 Residents - Bicycle: Score` = as.numeric(`Road Traffic Fatalities per 100,000 Residents - Bicycle: Score`),
  #   `Road Traffic Fatalities per 100,000 Residents - Pedestrian: Score` = as.numeric(`Road Traffic Fatalities per 100,000 Residents - Pedestrian: Score`)
  # ) %>%
  group_by(region) %>%
  summarise(
    center_long = mean(long, na.rm = TRUE),
    center_lat = mean(lat, na.rm = TRUE),
    auto_score = mean(`Road.Traffic.Fatalities.per.100.000.Residents...Auto..Score`, na.rm = TRUE),
    bicycle_score = mean(`Road.Traffic.Fatalities.per.100.000.Residents...Bicycle..Score`, na.rm = TRUE),
    pedestrian_score = mean(`Road.Traffic.Fatalities.per.100.000.Residents...Pedestrian..Score`, na.rm = TRUE),
    state_name = first(state)
  )

selected_state_data <- state_centers %>%
            filter(state_name == input$state) %>%
            select(center_long, center_lat, auto_score, bicycle_score, pedestrian_score)

        # Score to be displayed based on the selected transportation type
        transport_type_score <- switch(input$transportType,
            "Road.Traffic.Fatalities.per.100.000.Residents...Auto..Score" = selected_state_data$auto_score,
            "Road.Traffic.Fatalities.per.100.000.Residents...Bicycle..Score" = selected_state_data$bicycle_score,
            "Road.Traffic.Fatalities.per.100.000.Residents...Pedestrian..Score" = selected_state_data$pedestrian_score
        )
# Create a color palette for the activity score
pal <- colorNumeric(palette = "viridis", domain = state_centers[[input$transportType]])

output$stateMap <- renderLeaflet({
            if (nrow(selected_state_data) > 0) {
                leaflet() %>%
                    addTiles() %>%
                    addMarkers(
                        lng = selected_state_data$center_long, 
                        lat = selected_state_data$center_lat, 
                        popup = paste(input$state, input$transportType, "Score:", transport_type_score)
                    ) %>%
                    setView(lng = selected_state_data$center_long, lat = selected_state_data$center_lat, zoom = 6) 
            } else { 
                leaflet() %>%
                    addTiles() %>%
                    setView(lng = -96.9, lat = 37.8, zoom = 4) 
            } 
        })
    })} 

# Run the Shiny app
shinyApp(ui = ui, server = server)

```
